---
title: "Transportability Testing with mrtransport"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Transportability Testing with mrtransport}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mrtransport)
```

## What mrtransport does

Multi-ancestry Mendelian randomization estimates sometimes disagree across
populations. `mrtransport` provides two complementary tools:

1. **Scalar transportability testing**: Cochran's Q with post-hoc power
   analysis, sensitivity diagnostics, and a "transport" / "non-transport"
   verdict for single-trait MR estimates.

2. **Vector sheaf Q**: A multivariate Q statistic that detects effect vector
   rotation across populations at multi-trait loci -- heterogeneity patterns
   that per-trait scalar Q tests miss entirely.

This tutorial walks through both workflows.

---

## Part 1: Scalar transportability testing

### Case 1: HLA-DRB1 shared epitope to RA -- clear non-transport

The HLA-DRB1 shared epitope has dramatically different effects on rheumatoid
arthritis risk across ancestries (EUR OR~5.0, EAS OR~3.3, AFR OR~2.2). This
is a textbook non-transportable pair.

```{r case1}
result <- test_transport(
  betas = c(1.6, 1.2, 0.8, 1.0),
  ses   = c(0.04, 0.07, 0.12, 0.10),
  strata_names = c("EUR", "EAS", "AFR", "HISP")
)
cat(sprintf("Verdict:    %s\n", result$verdict))
cat(sprintf("Q = %.2f, df = %d, p = %.2e\n", result$Q, result$df, result$p))
cat(sprintf("I2 = %.1f%%, power = %.2f\n", result$I2 * 100, result$power))
cat(sprintf("Underpowered? %s\n", result$underpowered))
```

The Q test easily rejects homogeneity. High power confirms this is a
real signal, not a type I error.

### Case 2: SBP to Alzheimer's disease -- underpowered

Systolic blood pressure and Alzheimer's disease: the effects look
heterogeneous (I2~58%), but only 4 strata produce low power. The test
correctly returns "transport" (non-significant), but the `underpowered`
flag warns that this is honest uncertainty, not evidence of homogeneity.

```{r case2}
result_sbp <- test_transport(
  betas = c(0.10, -0.05, 0.15, 0.02),
  ses   = c(0.04, 0.05, 0.08, 0.10),
  strata_names = c("EUR_midlife", "EUR_latelife", "EAS", "AFR")
)
cat(sprintf("Verdict:    %s\n", result_sbp$verdict))
cat(sprintf("Q = %.2f, df = %d, p = %.4f\n", result_sbp$Q, result_sbp$df, result_sbp$p))
cat(sprintf("I2 = %.1f%%, power = %.2f\n", result_sbp$I2 * 100, result_sbp$power))
cat(sprintf("Underpowered? %s\n", result_sbp$underpowered))
```

How many strata would achieve 80% power at this I2 level?

```{r strata-needed}
k_needed <- strata_needed(I2 = result_sbp$I2, target_power = 0.80)
cat(sprintf("Strata needed for 80%% power: %s\n",
            if (is.null(k_needed)) ">50" else as.character(k_needed)))
```

### Case 3: Synthetic catalog -- benchmarking accuracy

Generate 20 transportable and 10 non-transportable pairs with known ground
truth, classify all, and check accuracy.

```{r case3}
catalog <- simulate_catalog(n_transport = 20, n_nontransport = 10, seed = 42)
correct <- 0L
for (pair in catalog) {
  r <- test_transport(
    betas = pair$betas,
    ses = pair$ses,
    strata_names = pair$strata_names
  )
  if (r$verdict == pair$ground_truth) correct <- correct + 1L
}
cat(sprintf("Accuracy: %d/%d (%.0f%%)\n", correct, length(catalog),
            100 * correct / length(catalog)))
```

### Case 4: Bring your own data -- with sensitivity analysis

Replace the vectors below with your own per-stratum MR estimates.

```{r case4}
my_result <- test_transport(
  betas = c(0.25, 0.23, 0.28),
  ses   = c(0.05, 0.06, 0.07),
  strata_names = c("EUR", "EAS", "AFR")
)
cat(sprintf("Verdict: %s (Q = %.2f, p = %.4f, power = %.2f)\n",
            my_result$verdict, my_result$Q, my_result$p, my_result$power))
```

**Alpha sweep**: how does the verdict change across significance thresholds?

```{r sweep}
sweep <- alpha_sweep(
  betas = c(0.25, 0.23, 0.28),
  ses   = c(0.05, 0.06, 0.07)
)
for (i in seq_len(nrow(sweep))) {
  cat(sprintf("  alpha=%.3f: %-15s (power=%.2f)\n",
              sweep$alpha[i], sweep$verdict[i], sweep$power[i]))
}
```

**Leave-one-out**: does the verdict depend on a single stratum?

```{r loo}
loo <- leave_one_out(
  betas = c(0.25, 0.23, 0.28),
  ses   = c(0.05, 0.06, 0.07),
  strata_names = c("EUR", "EAS", "AFR")
)
cat(sprintf("LOO stable: %s\n", loo$stable))
if (!loo$stable) {
  cat(sprintf("Flips when dropping: %s\n", paste(loo$flipped_strata, collapse = ", ")))
}
```

---

## Part 2: Vector sheaf Q (multi-trait loci)

Standard practice runs Cochran's Q on each trait separately, but this misses
a specific failure mode: **effect vector rotation**, where per-trait effect
magnitudes look similar but the joint direction of the effect vector changes
across ancestries.

The vector sheaf Q statistic tests for heterogeneity in the joint effect
vector. On a single trait (d = 1), it reduces algebraically to Cochran's Q.
On multiple traits (d > 1), it detects rotation that d independent scalar
tests cannot.

### Case 5: ACKR1/Duffy -- effect vector rotation across ancestries

The ACKR1/Duffy locus (rs2814778) affects multiple blood cell traits. The
Duffy-null allele, nearly fixed in sub-Saharan African populations due to
malaria selection, eliminates ACKR1-mediated chemokine scavenging. This
produces a coordinated multi-trait shift -- a rotation, not a uniform
scaling -- across ancestries.

```{r ackr1-data}
beta <- matrix(c(
   0.15,  0.08,  0.03,  0.12,  0.02,
   0.18,  0.10,  0.02,  0.14,  0.03,
  -0.25,  0.35,  0.20, -0.10,  0.15,
   0.12,  0.06,  0.04,  0.10,  0.01,
   0.08,  0.12,  0.06,  0.05,  0.04,
   0.14,  0.09,  0.03,  0.11,  0.02
), nrow = 6, byrow = TRUE)

se <- matrix(c(
  0.02, 0.02, 0.01, 0.02, 0.01,
  0.03, 0.03, 0.02, 0.03, 0.02,
  0.04, 0.04, 0.03, 0.04, 0.03,
  0.03, 0.03, 0.02, 0.03, 0.02,
  0.04, 0.04, 0.03, 0.04, 0.03,
  0.03, 0.03, 0.02, 0.03, 0.02
), nrow = 6, byrow = TRUE)

rownames(beta) <- rownames(se) <- c("EUR", "EAS", "AFR", "CSA", "AMR", "MID")
colnames(beta) <- colnames(se) <- c("Neutrophil", "Monocyte", "Basophil", "WBC", "Eosinophil")
```

Run the vector sheaf Q test:

```{r ackr1-vector}
result <- vector_sheaf_q_diagonal(beta, se)
cat(sprintf("Q_V = %.1f, df = %d, p = %.2e\n", result$Q_V, result$df, result$p))
```

Check if individual scalar Q tests on each trait detect this:

```{r ackr1-scalar}
d <- ncol(beta)
alpha_bonf <- 0.05 / d

for (j in seq_len(d)) {
  sq <- scalar_cochran_q(beta[, j], se[, j])
  sig <- if (sq$p < alpha_bonf) "*" else " "
  cat(sprintf("  %-12s  Q = %6.2f  p = %.4f  I2 = %.1f%%  %s\n",
              colnames(beta)[j], sq$Q, sq$p, sq$I2 * 100, sig))
}
```

The vector test detects coordinated heterogeneity that some individual scalar
tests miss. The rotation angle shows which ancestry pair diverges most:

```{r ackr1-rotation}
rot <- max_rotation(beta)
cat(sprintf("Maximum rotation: %.1f degrees (between %s and %s)\n",
            rot$theta_max,
            rownames(beta)[rot$pair[1]],
            rownames(beta)[rot$pair[2]]))
```

### Case 6: Vector Q vs scalar Q -- four-way classification

`classify_locus()` runs both tests and assigns one of four categories:

- **concordant_significant**: both vector Q and at least one scalar Q significant
- **concordant_null**: neither detects heterogeneity
- **rotation_only**: vector Q significant, no individual scalar Q significant (after Bonferroni)
- **component_only**: at least one scalar Q significant, but vector Q is not (rare)

```{r classify}
cl <- classify_locus(beta, se)
cat(sprintf("Category: %s\n", cl$category))
cat(sprintf("Vector Q_V: %.1f (p = %.2e)\n",
            cl$vector_result$Q_V, cl$vector_result$p))
cat(sprintf("Max rotation: %.1f degrees\n", cl$rotation$theta_max))
```

A "rotation_only" classification means the heterogeneity is invisible to
per-trait analysis. The vector test detects it because it accounts for the
coordinated shift in the joint effect vector.

### Case 7: Why trait correlations matter

Blood traits measured on the same individuals are correlated (shared
hematopoietic pathways). When you know the within-stratum trait correlations,
`vector_sheaf_q_correlated()` uses them to build a more accurate covariance
model.

```{r correlated}
diag_result <- vector_sheaf_q_diagonal(beta, se)

d <- ncol(beta)
rho <- 0.4
rho_matrix <- matrix(rho, nrow = d, ncol = d)
diag(rho_matrix) <- 1

corr_result <- vector_sheaf_q_correlated(beta, se, rho_matrix)

cat(sprintf("Diagonal Q_V:    %8.1f  (p = %.2e)\n",
            diag_result$Q_V, diag_result$p))
cat(sprintf("Correlated Q_V:  %8.1f  (p = %.2e)\n",
            corr_result$Q_V, corr_result$p))
cat(sprintf("Ratio Q_V / sum(scalar Q): %.2f\n",
            corr_result$Q_V / corr_result$sum_scalar_Q))
```

When traits are positively correlated and heterogeneity is rotation-like,
the correlated Q_V diverges upward from the sum of scalar Q tests. This
ratio > 1 is the signature of covariance-mediated heterogeneity that
per-trait tests cannot capture.

**Key invariant**: with diagonal covariance (rho = 0), Q_V equals the sum
of scalar Q tests exactly. The correlated version diverges only when
off-diagonal structure interacts with the heterogeneity pattern.

```{r invariant}
cat(sprintf("Diagonal: Q_V = %.4f, sum(scalar) = %.4f, difference = %.1e\n",
            diag_result$Q_V, diag_result$sum_scalar_Q,
            abs(diag_result$Q_V - diag_result$sum_scalar_Q)))
```

### Case 8: Bring your own multi-trait data

Replace the matrices below with your own per-ancestry, per-trait MR estimates.

```{r byod-vector}
my_beta <- matrix(c(
  0.25, 0.10,
  0.23, 0.12,
  0.28, 0.08
), nrow = 3, byrow = TRUE)

my_se <- matrix(c(
  0.05, 0.03,
  0.06, 0.04,
  0.07, 0.05
), nrow = 3, byrow = TRUE)

rownames(my_beta) <- rownames(my_se) <- c("EUR", "EAS", "AFR")
colnames(my_beta) <- colnames(my_se) <- c("LDL", "HDL")

vq <- vector_sheaf_q_diagonal(my_beta, my_se)
cat(sprintf("Vector Q_V = %.2f, df = %d, p = %.4f\n", vq$Q_V, vq$df, vq$p))

for (j in seq_len(ncol(my_beta))) {
  sq <- scalar_cochran_q(my_beta[, j], my_se[, j])
  cat(sprintf("  %s: Q = %.2f, p = %.4f, I2 = %.0f%%\n",
              colnames(my_beta)[j], sq$Q, sq$p, sq$I2 * 100))
}

cl <- classify_locus(my_beta, my_se)
cat(sprintf("Classification: %s\n", cl$category))
cat(sprintf("Max rotation: %.1f degrees\n", cl$rotation$theta_max))
```

With known trait correlations:

```{r byod-correlated}
rho <- matrix(c(1.0, -0.3,
               -0.3,  1.0), nrow = 2)

vq_corr <- vector_sheaf_q_correlated(my_beta, my_se, rho)
cat(sprintf("Correlated Q_V = %.2f (p = %.4f)\n", vq_corr$Q_V, vq_corr$p))
cat(sprintf("Q_V / sum(scalar): %.2f\n", vq_corr$Q_V / vq_corr$sum_scalar_Q))
```

---

## Interpretation guide

### Scalar transportability

| Verdict | Power | I2 | Interpretation |
|:--|:--|:--|:--|
| non-transport | high | high | Clear heterogeneity; effect does not transport |
| transport | high | low | Homogeneous; effect likely transports |
| transport | low | high | **Underpowered**: honest uncertainty, get more strata |
| transport | low | low | Insufficient signal either way |

### Vector Q classification

| Q_V significant? | Any scalar Q significant? | Category | Interpretation |
|:-:|:-:|:--|:--|
| Yes | Yes | concordant_significant | Strong heterogeneity in both joint and marginal tests |
| No | No | concordant_null | No detectable heterogeneity |
| Yes | No | rotation_only | Effect vector rotates; no single trait is heterogeneous enough alone |
| No | Yes | component_only | Anti-correlated heterogeneity cancels in the joint test (rare) |
