---
title: "Detecting Effect Vector Rotation with sheafq"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Detecting Effect Vector Rotation with sheafq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(sheafq)
```

## What sheafq does

Multi-ancestry Mendelian randomization estimates sometimes disagree across
populations. Standard practice runs Cochran's Q on each trait separately, but
this misses a specific failure mode: **effect vector rotation**, where per-trait
effect magnitudes look similar but the joint direction of the effect vector
changes across ancestries.

`sheafq` implements the vector-valued sheaf Q statistic, which tests for
heterogeneity in the joint effect vector. On a single trait (d = 1), it reduces
algebraically to Cochran's Q. On multiple traits (d > 1), it detects rotation
that d independent scalar tests cannot.

This tutorial walks through four use cases:

1. ACKR1/Duffy: a textbook rotation locus
2. Comparing vector Q to per-trait scalar Q
3. The role of trait correlations
4. Bring your own data

## Case 1: ACKR1/Duffy — effect vector rotation across ancestries

The ACKR1/Duffy locus (rs2814778) affects multiple blood cell traits. The
Duffy-null allele, nearly fixed in sub-Saharan African populations due to
malaria selection, eliminates ACKR1-mediated chemokine scavenging. This
produces a coordinated multi-trait shift — a rotation, not a uniform scaling —
across ancestries.

```{r ackr1-data}
# Per-ancestry effect estimates at ACKR1 for 5 blood traits
# Rows: EUR, EAS, AFR, CSA, AMR, MID
# Columns: neutrophil, monocyte, basophil, WBC, eosinophil
beta <- matrix(c(
   0.15,  0.08,  0.03,  0.12,  0.02,
   0.18,  0.10,  0.02,  0.14,  0.03,
  -0.25,  0.35,  0.20, -0.10,  0.15,
   0.12,  0.06,  0.04,  0.10,  0.01,
   0.08,  0.12,  0.06,  0.05,  0.04,
   0.14,  0.09,  0.03,  0.11,  0.02
), nrow = 6, byrow = TRUE)

se <- matrix(c(
  0.02, 0.02, 0.01, 0.02, 0.01,
  0.03, 0.03, 0.02, 0.03, 0.02,
  0.04, 0.04, 0.03, 0.04, 0.03,
  0.03, 0.03, 0.02, 0.03, 0.02,
  0.04, 0.04, 0.03, 0.04, 0.03,
  0.03, 0.03, 0.02, 0.03, 0.02
), nrow = 6, byrow = TRUE)

rownames(beta) <- rownames(se) <- c("EUR", "EAS", "AFR", "CSA", "AMR", "MID")
colnames(beta) <- colnames(se) <- c("Neutrophil", "Monocyte", "Basophil", "WBC", "Eosinophil")
```

Run the vector sheaf Q test:

```{r ackr1-vector}
result <- vector_sheaf_q_diagonal(beta, se)
cat(sprintf("Q_V = %.1f, df = %d, p = %.2e\n", result$Q_V, result$df, result$p))
```

Now check if individual scalar Q tests on each trait detect this:

```{r ackr1-scalar}
d <- ncol(beta)
alpha_bonf <- 0.05 / d

for (j in seq_len(d)) {
  sq <- scalar_cochran_q(beta[, j], se[, j])
  sig <- if (sq$p < alpha_bonf) "*" else " "
  cat(sprintf("  %-12s  Q = %6.2f  p = %.4f  I2 = %.1f%%  %s\n",
              colnames(beta)[j], sq$Q, sq$p, sq$I2 * 100, sig))
}
```

The vector test detects coordinated heterogeneity that some individual scalar
tests miss. The rotation angle shows which ancestry pair diverges most:

```{r ackr1-rotation}
rot <- max_rotation(beta)
cat(sprintf("Maximum rotation: %.1f degrees (between %s and %s)\n",
            rot$theta_max,
            rownames(beta)[rot$pair[1]],
            rownames(beta)[rot$pair[2]]))
```

## Case 2: Vector Q vs scalar Q — when they agree and when they don't

`classify_locus()` runs both tests and assigns one of four categories:

- **concordant_significant**: both vector Q and at least one scalar Q significant
- **concordant_null**: neither detects heterogeneity
- **rotation_only**: vector Q significant, no individual scalar Q significant (after Bonferroni)
- **component_only**: at least one scalar Q significant, but vector Q is not (rare)

```{r classify}
result <- classify_locus(beta, se)
cat(sprintf("Category: %s\n", result$category))
cat(sprintf("Vector Q_V: %.1f (p = %.2e)\n",
            result$vector_result$Q_V, result$vector_result$p))
cat(sprintf("Max rotation: %.1f degrees\n", result$rotation$theta_max))
```

A "rotation_only" classification means the heterogeneity is invisible to
per-trait analysis. The vector test detects it because it accounts for the
coordinated shift in the joint effect vector.

## Case 3: Why trait correlations matter

Blood traits measured on the same individuals are correlated (shared hematopoietic
pathways). When you know the within-stratum trait correlations, `vector_sheaf_q_correlated()`
uses them to build a more accurate covariance model.

```{r correlated}
# With diagonal covariance (assumes independent traits):
diag_result <- vector_sheaf_q_diagonal(beta, se)

# With a compound-symmetric correlation (rho = 0.4 across all traits):
d <- ncol(beta)
rho <- 0.4
rho_matrix <- matrix(rho, nrow = d, ncol = d)
diag(rho_matrix) <- 1

corr_result <- vector_sheaf_q_correlated(beta, se, rho_matrix)

cat(sprintf("Diagonal Q_V:    %8.1f  (p = %.2e)\n",
            diag_result$Q_V, diag_result$p))
cat(sprintf("Correlated Q_V:  %8.1f  (p = %.2e)\n",
            corr_result$Q_V, corr_result$p))
cat(sprintf("Ratio Q_V / sum(scalar Q): %.2f\n",
            corr_result$Q_V / corr_result$sum_scalar_Q))
```

When traits are positively correlated and heterogeneity is
rotation-like, the correlated Q_V diverges upward from the sum of scalar Q
tests. This ratio > 1 is the signature of covariance-mediated heterogeneity
that per-trait tests cannot capture.

**Key invariant**: with diagonal covariance (rho = 0), Q_V equals the sum of
scalar Q tests exactly. The correlated version diverges only when off-diagonal
structure interacts with the heterogeneity pattern.

```{r invariant}
cat(sprintf("Diagonal: Q_V = %.4f, sum(scalar) = %.4f, difference = %.1e\n",
            diag_result$Q_V, diag_result$sum_scalar_Q,
            abs(diag_result$Q_V - diag_result$sum_scalar_Q)))
```

## Case 4: Bring your own data

Replace the matrices below with your own per-ancestry, per-trait MR estimates.

```{r byod}
# Your data: K ancestries x d traits
my_beta <- matrix(c(
  0.25, 0.10,   # Ancestry 1
  0.23, 0.12,   # Ancestry 2
  0.28, 0.08    # Ancestry 3
), nrow = 3, byrow = TRUE)

my_se <- matrix(c(
  0.05, 0.03,
  0.06, 0.04,
  0.07, 0.05
), nrow = 3, byrow = TRUE)

rownames(my_beta) <- rownames(my_se) <- c("EUR", "EAS", "AFR")
colnames(my_beta) <- colnames(my_se) <- c("LDL", "HDL")

# Run vector Q
vq <- vector_sheaf_q_diagonal(my_beta, my_se)
cat(sprintf("Vector Q_V = %.2f, df = %d, p = %.4f\n", vq$Q_V, vq$df, vq$p))

# Run per-trait scalar Q for comparison
for (j in seq_len(ncol(my_beta))) {
  sq <- scalar_cochran_q(my_beta[, j], my_se[, j])
  cat(sprintf("  %s: Q = %.2f, p = %.4f, I2 = %.0f%%\n",
              colnames(my_beta)[j], sq$Q, sq$p, sq$I2 * 100))
}

# Classify
cl <- classify_locus(my_beta, my_se)
cat(sprintf("Classification: %s\n", cl$category))
cat(sprintf("Max rotation: %.1f degrees\n", cl$rotation$theta_max))
```

### With known trait correlations

If your traits were measured in the same sample and you have the phenotypic
correlation matrix:

```{r byod-correlated}
rho <- matrix(c(1.0, -0.3,
               -0.3,  1.0), nrow = 2)

vq_corr <- vector_sheaf_q_correlated(my_beta, my_se, rho)
cat(sprintf("Correlated Q_V = %.2f (p = %.4f)\n", vq_corr$Q_V, vq_corr$p))
cat(sprintf("Q_V / sum(scalar): %.2f\n", vq_corr$Q_V / vq_corr$sum_scalar_Q))
```

## Interpretation guide

| Q_V significant? | Any scalar Q significant? | Category | Interpretation |
|:-:|:-:|:--|:--|
| Yes | Yes | concordant_significant | Strong heterogeneity visible in both joint and marginal tests |
| No | No | concordant_null | No detectable heterogeneity |
| Yes | No | rotation_only | Effect vector rotates across ancestries, but no single trait is heterogeneous enough to detect alone |
| No | Yes | component_only | Rare: anti-correlated heterogeneity cancels in the joint test |

The "rotation_only" category is the reason sheafq exists. These loci would be
missed entirely by standard per-trait Cochran's Q analysis.
